I don't know, are you sure we want to do this?\\ Sea level adaptation decisions under uncertainty

### 2016-12-30 submission/paper.tex
Added text to Bergen adaptation section. Minor corrections throughout

\begin{abstract}
Sea level rise has serious consequences for harbor infrastructure, storm drains and sewer systems, and many other issues. Adapting to sea level rise requires comparing different possible adaptation strategies, comparing the cost of different actions (including no action), and assessing at what point in time the chosen strategy should be implemented. All these decisions must be made under considerable uncertainty--in the amount of sea level rise, in the cost of adaptation actions, and in the cost of no action. We develop two illustrative examples: for Bergen on Norway's west coast and for Esbjerg on the west coast of Denmark. Different components of uncertainty are visualized. We show that failing to take uncertainty into account can increase the projected damage costs by an order of magnitude.
\end{abstract}

@@

As adaptation decision-making is an ongoing process of weighing and choosing which measures should be taken at which moment in time \citep{Hallegatte&2012} adaptive planning methods need to support decisions in the short term, while considering long-term developments. Challenges of adaptation decision-making under uncertainty relate to the incorporation of spatial, inter-temporal and flexibility aspects of adaptation priorities \citep{FankhauserSoare2013}, and the linkage with specific characteristics of sectors and contexts \citep{BisaroSwartHinkel2016, HinkelBisaro2016}. Several economic decision support tools and methods exist for adaptation assessment under uncertainty \citep[e.g.][]{Chambwera&2014, WilbyDessai2010, WalkerHaasnootKwakkel2013}. However, \citet{Watkiss&2015} conclude that these tools are very resource intensive and complex in the context of long-term adaptation investment decisions and call for the development of ``light touch'' approaches to better support local adaptation making.

In this paper we employ light touch decision tools to demonstrate the importance of combining projections of sea level rise and flood damages alongside a detailed quantification of both hydrologic and economic uncertainties in the context of real-life decision-problems experienced by stakeholders and authorities in two northern European cities, Bergen in Norway and Esbjerg in Denmark, see Figure~\ref{fig:CityMaps}. Based on communications with local end-users we highlight the value of taking into account uncertainty through two simplified and complementary case studies, where in the first one  planners want to know how early they should implement costly adaptation measures, whereas in the second case the aim is to highlight the risk of flooding in coastal areas, e.g. in order to prioritize future adaptation actions and investments. In both cases we show that embracing the uncertainties derived from economic and hydrologic models is absolutely crucial in order to answer the question of "are we sure we want to do this?".

@@

The municipality of Bergen has, in cooperation with private actors, analyzed several possible adaptation measures against sea level rise. The measures range from an outer barrier that would protect the entire metropolitan area to various protection measures of limited areas in the inner harbor \citep{bergenreport}. While the viability of the constructions and the associated construction costs have been carefully analyzed, the optimal timing of potential adaptation measures and the effects of the associated uncertainties have yet to be investigated. We perform such an analysis where we consider uncertainty in projected sea level rise, damage costs and the effects of sea level rise on changes in damage costs.

Esbjerg, on the southwest coast of Jutland, is the fifth-largest city in Denmark and the largest urban area in the region. The city hosts one of the largest harbours  in Denmark, which serves as a focal point for offshore activities in the North Sea, including the continued development of offshore wind power and extensive activities related to the extraction of oil and gas. As a result critical infrastructures and commercial buildings figure prominently in the coastal zone. Esbjerg is frequently subject to substantial storms and storm surges, causing severe flooding of the harbour and the city. The highest ever since 1874 was recorded in 1981, where the harbour was completely flooded and the water level reached a record-high 433 cm above the norm, causing massive economic losses.  More generally, storm surges causing water levels in Esbjerg to rise to between 2 and 3 metres have quadrupled over the last four decades according to local records, whereas half of the most severe events have taken place since 1975.

@@

The remainder of the paper is organized as follows. Section \ref{sealevelproj} describes our approach to projecting sea level. In sections \ref{decision_tools_PartI} and \ref{decision_tools_PartII} we describe the type of decision problems that we are going to attack. We apply these tools to sea level projections for Bergen, Norway, and for Esbjerg, Denmark in section \ref{cases}. In section \ref{unc} we demonstrate the consequences of ignoring the uncertainty in the projections, and the paper is closed with a summary and discussion in section \ref{disc}..

@@

\subsection{Global sea level}
Most climate models do not explicitly provide sea level as an output of the calculations. Rather, the IPCC AR5 report \citep[ch.~13]{ipcc} combines the heat expansion of the ocean with temperature forced models for glacial melt, Greenland ice melt, and Antarctic ice melt and with land rise due to rebound from the last ice age. Judging from the supplementary material to \citet[ch.~13]{ipcc}, the uncertainty assessment is only based on the spread of the ensemble of temperature projections, not on the additional uncertainty in the ice models used.

We will instead use the empirical approach of Rahmstorf and collaborators \citep{Rahmstorf07,Rahmstorf11}, employing the statistical modeling of \citet{Bolin2014a} to relate global annual mean temperature anomalies to global mean sea level anomalies.

We then apply the estimated historical relationship to projected temperatures from the CMIP5 experiment \citep{cmip5} to obtain projected global annual mean sea level, taking into account the uncertainty in the statistical model as well as the spread of the temperature projection ensemble (see subsection \ref{unc_ass}). 
For the i'th temperature projection $T_t^i$ we estimate the corresponding global mean sea level as
\begin{linenomath*}
\[H_t^{gl,i} = \int\limits_{{t_0}}^t {{\hat a} (T_u^i - {{\hat T}_0}} )du + {\varsigma _t},\]
\end{linenomath*}
where ${\hat a}$ and ${\hat T}_0$ are regression parameters of observed global sea level  rise on observed global temperature and $\varsigma_t$ the integrated time series regression error.

@@

There is an increasing need for more detailed economic analysis, including simple methods and tools for assessment of options, especially since as Downing (2012) recognizes adaptation is moving from theory to practice and practitioners try to deal with how to begin adapting. This leads to an increasing need and interest in the appraisal of options.

Several economic decision support tools and methods exist for adaptation assessment under uncertainty. Robust decision-making approaches are able to better incorporate uncertainty and a broad range of climate scenarios to capture as much of the uncertainty on future climates as possible (Dittrich et al. 2016). These approaches can be classified according to a science-first or policy-first approach. The former has a €``predict-then-act''€™ foundation, which starts with for example probabilistic climate projections and impact assessments, not linked to any specific adaptation choices (Jones et al., 2014). The latter starts out with the formulated adaptation plans and not impacts and their functioning is tested against different future projections (Dittrich et al. 2016).

@@

\subsection{Limitations of the decision framework}
The main limitation of this light touch decision framework is that we have significantly simplified the assessment of the effect of sea level rise on the damage costs. In particular, the linear extrapolation of the results reported in \citet{Hallegatte&2013} might provide a conservative estimate of the effect of extreme sea level rise. However, with only two data points, extrapolation approaces such as a power law or exponential growth seem unfeasible.

@@

@@

Figure \ref{fig:bergenobs} shows uncorrected and corrected Bergen sea level data, and the relationship between the corrected Bergen data and the global sea level data. The glacial isostatic adjustment is 0.26 (0.07) cm/yr. The time series regression uses an ARMA(1,1)-model \citep{boxjenkins}, with AR parameter 0.82 (0.13), and MA parameter --0.61 (0.17). The regression slope is 1.30 (0.12).

@@

Figure~\ref{fig:BergenDamageDist} shows the histogram of observed annual damage costs for Bergen and the associated Burr distribution. The parameter estimates are $\hat{\alpha} = 7.84$ (3.6), $\hat{\gamma} = 0.40$ (0.04) and $\hat{\theta} = 0.007$ (0.01). \citet{bergenreport} discuss several different adaptation options for Bergen. In Figure~\ref{fig:TotalDamageBergen} we consider the optimal timing of an adaptation option that includes two inner barriers at V{\aa}gen and Damg{\aa}rdssundet, that is, one on each side of central Bergen. The combined construction cost of the two barriers for a protection against 75 cm sea level rise is estimated at 1.13 billion NOK (2015 level).

@@

\begin{table}
\centering
\begin{tabular}[]{c | c| c c c}
\hline
&&&RCP 8.5&\\
Storm surge&No sea level rise&5th percentile&median&95th percentile\\
\hline
RP 20&388 cm&431 cm&464 cm& 500 cm\\
RP100&405 cm&448 cm& 481 cm& 517 cm\\
\hline
\end{tabular}
\label{tab:stormsurge}
\caption{Storm surge water levels \citep{sealevel2012} with a return period of 20 years (RP20) and 100 years (RP100) with no sea level rise  and with sea level rise corresponding to RCP 8.5 (5th percentile, median, and 95th percentile).}
\end{table}

@@

\subsection{Damage costs}
A simplistic analysis of projected damage costs at the year 2100, not taking into account the uncertainty, would use  the median  historical damage cost multiplied by  the median damage effect factor at 2100 at the median sea level rise projected for 2100. This yields a damage cost of NOK 335 million (the grey vertical line in Figure \ref{fig:costunc}. Similar results obtain when allowing sea level or effect factor to vary, holding the other quantities at the median (yellow and purple dots on top of Figure \ref{fig:costunc}). However, allowing only the damage cost to vary yields a median cost of 3.84 billion (green dot on top of Figure \ref{fig:costunc}). The appropriate uncertainty analysis for our model should draw each of sea level, effect factor and damage cost at random from their 2100 distributions. This corresponds to a median cost of 3.00 billion NOK, a factor of 9 higher than the simplistic value. Over 99\% of the costs in our model are higher than the simplistic median.

>>>>

\begin{abstract}
Sea level rise has serious consequences for harbor infrastructure, storm drains and sewer systems, and many other issues. Adapting to sea level rise requires comparing different possible adaptation strategies, comparing the cost of different actions (including no action), and assessing at what point in time the chosen strategy should be implemented. All these decisions must be made under considerable uncertainty--in the amount of sea level rise, in the cost of adaptation actions, and in the cost of no action. We develop two illustrative examples: for Bergen on Norway's west coast and for Esbjerg on the west coast of Denmark. Different components of uncertainty are visualized. We show that failing to take uncertainty into account can result in the median projected damage costs being an order of magnitude smaller.
\end{abstract}

@@

As adaptation decision-making is an ongoing process of weighing and choosing which measures should be taken at which moment in time \citep{Hallegatte&2012}, adaptive planning methods need to support decisions in the short term, while considering long-term developments. Challenges of adaptation decision-making under uncertainty relate to the incorporation of spatial, inter-temporal and flexibility aspects of adaptation priorities \citep{FankhauserSoare2013}, and the linkage with specific characteristics of sectors and contexts \citep{BisaroSwartHinkel2016, HinkelBisaro2016}. Several economic decision support tools and methods exist for adaptation assessment under uncertainty \citep[e.g.][]{Chambwera&2014, WilbyDessai2010, WalkerHaasnootKwakkel2013}. However, \citet{Watkiss&2015} conclude that these tools are very resource intensive and complex in the context of long-term adaptation investment decisions and call for the development of ``light touch'' approaches to better support local adaptation making.

In this paper we employ light touch decision tools to demonstrate the importance of combining projections of sea level rise and flood damages alongside a detailed quantification of both hydrologic and economic uncertainties in the context of real-life decision-problems experienced by stakeholders and authorities in two northern European cities, Bergen in Norway and Esbjerg in Denmark, see Figure~\ref{fig:CityMaps}. Based on communications with local end-users we highlight the value of taking into account uncertainty through two simplified and complementary case studies, where in the first one  planners want to know how early they should implement costly adaptation measures, whereas in the second case the aim is to highlight the risk of flooding in coastal areas, e.g. in order to prioritize future adaptation actions and investments. In both cases we show that embracing the uncertainties derived from economic and hydrologic models is absolutely crucial in order to answer the question of ``are we sure we want to do this?''

@@

The municipality of Bergen has, in cooperation with private actors, analyzed several possible adaptation measures against sea level rise. The measures range from an outer barrier that would protect the entire metropolitan area to various protection measures of limited areas in the inner harbor \citep{bergenreport}. While the viability of the constructions and the associated construction costs have been carefully analyzed, the optimal timing of potential adaptation measures and the effects of the associated uncertainties have yet to be investigated. We perform such an analysis where we consider uncertainty in projected sea level rise, damage costs, and the effect of sea level rise on changes in damage costs.

Esbjerg, on the southwest coast of Jutland, is the fifth-largest city in Denmark and the largest urban area in the region. The city hosts one of the largest harbours  in Denmark, which serves as a focal point for offshore activities in the North Sea, including the continued development of offshore wind power and extensive activities related to the extraction of oil and gas. As a result critical infrastructures and commercial buildings figure prominently in the coastal zone. Esbjerg is frequently subject to substantial storms and storm surges, causing severe flooding of the harbour and the city. The highest since records began in 1874 was recorded in 1981, where the harbour was completely flooded and the water level reached  433 cm above the norm, causing massive economic losses.  More generally, storm surges causing water levels in Esbjerg to rise to between 2 and 3 metres have quadrupled over the last four decades according to local records, whereas half of the most severe events have taken place since 1975.

@@

The remainder of the paper is organized as follows. Section \ref{sealevelproj} describes our approach to projecting sea level. In sections \ref{decision_tools_PartI} and \ref{decision_tools_PartII} we describe the type of decision problems that we are going to attack. We apply these tools to sea level projections for Bergen, Norway, and for Esbjerg, Denmark in section \ref{cases}. In section \ref{unc} we demonstrate the consequences of ignoring the uncertainty in the projections, and the paper is closed with a summary and discussion in section \ref{disc}.

@@

\subsection{Global sea level}
Most climate models do not explicitly provide sea level as an output of the calculations. Rather, the IPCC AR5 report \citep[ch.~13]{ipcc} combines the heat expansion of the ocean with temperature forced models for glacial melt, Greenland ice melt, and Antarctic ice melt and with land rise due to rebound from the last ice age and other tectonic effects. Judging from the supplementary material to \citet[ch.~13]{ipcc}, the uncertainty assessment is only based on the spread of the ensemble of temperature projections, not on the additional uncertainty in the ice models used.

We will instead use the empirical approach of Rahmstorf and collaborators \citep{Rahmstorf07,Rahmstorf11}, employing the statistical modeling of \citet{Bolin2014a} to relate global annual mean temperature anomalies to global mean sea level anomalies. 
We then apply the estimated historical relationship to projected temperatures from the CMIP5 experiment \citep{cmip5} to obtain projected global annual mean sea level, taking into account the uncertainty in the statistical model as well as the spread of the temperature projection ensemble (see subsection \ref{unc_ass}). 
For the i'th temperature projection $T_t^i$ we estimate the corresponding global mean sea level as
\begin{linenomath*}
\[H_t^{gl,i} = \int\limits_{{t_0}}^t {{\hat a} (T_u^i - {{\hat T}_0}} )du + {\varsigma _t},\]
\end{linenomath*}
where ${\hat a}$ and ${\hat T}_0$ are regression parameters of observed global sea level  rise on observed global temperature and $\varsigma_t$ the integrated time series regression error.

@@

There is an increasing need for more detailed economic analysis, including simple methods and tools for assessment of options, especially since as Downing (2012) recognizes, adaptation is moving from theory to practice, and practitioners try to deal with how to begin adapting. This leads to an increasing need and interest in the appraisal of options.

Several economic decision support tools and methods exist for adaptation assessment under uncertainty. Robust decision-making approaches are able to better incorporate uncertainty and a broad range of climate scenarios to capture as much of the uncertainty on future climates as possible (Dittrich et al. 2016). These approaches can be classified according to a science-first or policy-first approach. The former has a ``predict-then-act'' foundation, which starts with climate projections and impact assessments, not linked to any specific adaptation choices (Jones et al., 2014). The latter starts out with the formulated adaptation plans and not impacts, and their functioning is tested against different future projections (Dittrich et al. 2016).

@@

\subsection{Limitations of the decision framework}
The main limitation of this light touch decision framework is that we have significantly simplified the assessment of the effect of sea level rise on the damage costs. In particular, the linear extrapolation of the results reported in \citet{Hallegatte&2013} might provide a conservative estimate of the effect of extreme sea level rise. However, with only two data points, extrapolation approaches such as a power law or exponential growth seem difficult to justify\`.

@@

Storm surge data for Esbjerg ate obtained from the Danish Coastal Authority \citep{sealevel2012}.

@@

Figure \ref{fig:bergenobs} shows uncorrected and corrected Bergen sea level data, and the relationship between the corrected Bergen data and the global sea level data. The glacial isostatic adjustment is 0.26 (standard error 0.07) cm/yr. The time series regression uses an ARMA(1,1)-model \citep{boxjenkins}, with AR parameter 0.82 (0.13), and MA parameter --0.61 (0.17). The regression slope is 1.30 (0.12).

@@

Figure~\ref{fig:BergenDamageDist} shows the histogram of observed annual damage costs for Bergen and the associated Burr distribution. The parameter estimates are $\hat{\alpha} = 7.84$ (3.6), $\hat{\gamma} = 0.40$ (0.04) and $\hat{\theta} = 0.007$ (0.01). \citet{bergenreport} discuss several different adaptation options for Bergen. In Figure~\ref{fig:TotalDamageBergen} we consider the optimal timing of an adaptation option that includes two inner barriers at V{\aa}gen and Damg{\aa}rdssundet, that is, one on each side of central Bergen. The combined construction cost of the two barriers for a protection against 75 cm sea level rise is estimated at 1.13 billion NOK (2015 level)\footnote{100 NOK is about 11 EUR.}.

Applying the methodology from section \ref{decision_tools_PartI} we find that the optimal time of building the barriers is in 2046 (Figure \ref{fig:TotalDamageBergen}), and that by the year 2100 this decision will on average save about 1/3 of the median damage costs without adaptation (Figure \ref{fig:NoAction}). At the high end (or the 95th percentile of the damage distribution) adaptation saves even more, about 43\%.

@@

\begin{table}

\begin{center}
\begin{tabular}[]{c | c | c c c}

&&&RCP 8.5&\\
Storm surge&No sea level rise&5th percentile&median&95th percentile\\
\hline
RP 20&388 cm&431 cm&464 cm& 500 cm\\
RP100&405 cm&448 cm& 481 cm& 517 cm\\
\hline
\end{tabular}
\end{center}
\caption{Storm surge water levels \citep{sealevel2012} with a return period of 20 years (RP20) and 100 years (RP100) with no sea level rise  and with sea level rise corresponding to RCP 8.5 (5th percentile, median, and 95th percentile).}
\label{tab:stormsurge}
\end{table}

@@

\subsection{Damage costs}
A simplistic analysis of projected damage costs at the year 2100, not taking into account the uncertainty, would use  the median  historical damage cost multiplied by  the median damage effect factor at 2100 at the median sea level rise projected for 2100. This yields a damage cost of NOK 338 million (the grey vertical line in Figure \ref{fig:costunc}). Similar results obtain when allowing sea level or effect factor to vary, holding the other quantities at the median (yellow and purple dots on top of Figure \ref{fig:costunc}). However, allowing only the damage cost to vary yields a median cost of 3.85 billion (green dot on top of Figure \ref{fig:costunc}). The appropriate uncertainty analysis for our model should draw each of sea level, effect factor and damage cost at random from their 2100 distributions. This corresponds to a median cost of 3.15 billion NOK, over 9 times higher than the simplistic value. Over 99\% of the costs in our simulation are higher than the simplistic median.
